---
title: "Reproducible Research: Peer Assessment 1"

---


## Loading and preprocessing the data
```{r}
  data <- read.csv("activity.csv", na.strings = "NA")
  data$date <- as.Date(data$date)
  head(data)
```
## What is mean total number of steps taken per day?
```{r warning=FALSE}
  ### USE data.table AND ggplot2 PACKAGES
    library(data.table)
    library(ggplot2)
```  
```{r results='hide'} 
  ### AGGREGATE THE TOTAL STEPS GROUPED BY DATE
	  data <- data.table(data)
  	data[, total_steps:=sum(steps, na.rm=TRUE), by=date]
  	aggregate <- unique(data[,c('date','total_steps')])
```
```{r warning=FALSE, message = FALSE}
  ### VIEW SAMPLE DATA
    head(aggregate)
  ### DRAW A BARPLOT
    plot <- ggplot(data = aggregate, aes(x=date, y=total_steps)) 
    plot <- plot + xlab("Date") + ylab("Total Number of Steps") + ggtitle("Number of Steps on Daily Basis") 
    plot <- plot + geom_bar(stat='identity', width=0.75, position=position_dodge(0.7)) + geom_point()
    plot <- plot + geom_smooth() + geom_line(lwd=1, col = "red3",cex=0.5)
  ### ADD THEME
    plot <- plot + theme(plot.title = element_text(color="#666666", face="bold", size=22)) 
    plot <- plot + theme(axis.title = element_text(color="#666666", face="bold", size=17))     
```
```{r results='hide', echo = FALSE} 
    plot
    with(plot,dev.copy(png, 'plot1.png'))
    dev.off()
```
```{r eval = FALSE} 
  ### COPY PLOT TO PLOT1.PNG
    with(plot,dev.copy(png, 'plot1.png'))
  ### CLOSE THE GRAPHICS DEVICE
    dev.off()
```
```{r results='hide'}
  ### CALCULATE MEAN AND MEDIAN OF THE TOTAL NUMBER OF STEPS TAKEN EACH DAY
    data[, mean:=mean(steps, na.rm = TRUE), by=date]
    data[, median:=as.numeric(median(steps, na.rm = TRUE)), by=date]
```
```{r}
  ### TAKE UNIQUE FOR DATE, TOTAL_STEPS, MEAN AND MEDIAN
    summary <- unique(data[,c('date','total_steps','mean','median')])
    summary
```
## What is the average daily activity pattern?
```{r results='hide'} 
  ### AGGREGATE THE TOTAL STEPS GROUPED BY INTERVAL
    data <- data.table(data)
  	data[, total_steps_per_interval:=sum(steps, na.rm=TRUE), by=interval]
  	aggregate_per_interval <- unique(data[,c('interval','total_steps_per_interval')])
```
```{r message = FALSE}
  ### DRAW A TIME SERIES PLOT SPLITTED ACROSS VARIOUS INTERVALS
    plot <- ggplot(data = aggregate_per_interval, aes(x=interval, y=total_steps_per_interval)) 
    plot <- plot + xlab("Interval") + ylab("Total Number of Steps") + ggtitle("Number of Steps Grouped by Interval") 
    plot <- plot + geom_line(color = "#FF0000")
    plot <- plot + theme(plot.title = element_text(color="#666666", face="bold", size=22)) 
    plot <- plot + theme(axis.title = element_text(color="#666666", face="bold", size=17))     
```
```{r results='hide', echo = FALSE} 
    plot
    with(plot,dev.copy(png, 'plot2.png'))
    dev.off()
```
```{r eval = FALSE} 
  ### COPY PLOT TO PLOT2.PNG
    with(plot,dev.copy(png, 'plot2.png'))
  ### CLOSE THE GRAPHICS DEVICE
    dev.off()
```
```{r}
  ### COMPUTE THE INTERVAL HAVING MAXIMUM NUMBER OF STEPS  
    aggregate_per_interval[which.max(aggregate_per_interval$total_steps_per_interval)]
```
## Imputing missing values
### Imputed the NAs with the mean of total number of steps for the particular day. Kindly note that, nothing has been imputed for the days where there were no steps taken for any day
```{r warning = FALSE}
  ### CREATE A NEW DATA FRAME    
    new_dataset <- data[,c('date', 'steps', 'interval','mean')]
  ### GET THE TOTAL NUMBER OF RECORDS WITH "NA"
    nrow(new_dataset[is.na(new_dataset$steps) == TRUE])
  ### IMPUT "NA" WITH MEANS FOR THE PARTICULAR DAY (IGNORING NaN)
    new_dataset$steps[is.na(new_dataset$steps)] <- new_dataset$mean
  ### NUMBER OF RECORDS LEFT AS "NA" (DUE TO THE REASON "ZERO STEPS IN ENTIRE DAY")
    nrow(new_dataset[is.na(new_dataset$steps) == TRUE])
### AGGREGATE THE TOTAL STEPS IN NEW DATA SET GROUPED BY DATE
    new_dataset <- data.table(new_dataset)
  	new_dataset[, total_steps:=sum(steps, na.rm=TRUE), by=date]
  	aggregate_new <- unique(new_dataset[,c('date','total_steps')])
```
```{r warning=FALSE, message = FALSE}
  ### DRAW A BARPLOT
    plot <- ggplot(data = aggregate_new, aes(x=date, y=total_steps)) 
    plot <- plot + xlab("Date") + ylab("Total Number of Steps") + ggtitle("Number of Steps on Daily Basis (After Imputing) ") 
    plot <- plot + geom_bar(stat='identity', width=0.75, position=position_dodge(0.7)) + geom_point()
    plot <- plot + geom_smooth() + geom_line(lwd=1, col = "blue",cex=0.5)
  ### ADD THEME
    plot <- plot + theme(plot.title = element_text(color="#666666", face="bold", size=18)) 
    plot <- plot + theme(axis.title = element_text(color="#666666", face="bold", size=15))     
```
```{r results='hide', echo = FALSE} 
    plot
    with(plot,dev.copy(png, 'plot3.png'))
    dev.off()
```
```{r eval = FALSE} 
  ### COPY PLOT TO PLOT3.PNG
    with(plot,dev.copy(png, 'plot3.png'))
  ### CLOSE THE GRAPHICS DEVICE
    dev.off()  
```
```{r results='hide'}
  ### CALCULATE MEAN AND MEDIAN OF THE TOTAL NUMBER OF STEPS TAKEN EACH DAY ON IMPUTED DATA SET
    new_dataset[, mean:=mean(steps, na.rm = TRUE), by=date]
    new_dataset[, median:=as.numeric(median(steps, na.rm = TRUE)), by=date]
```
```{r}
  ### TAKE UNIQUE FOR DATE, TOTAL_STEPS, MEAN AND MEDIAN ON IMPUTED DATA SET
    summary_new <- unique(new_dataset[,c('date','total_steps','mean','median')])
    summary_new
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
  ### ADD A NEW FACTOR VARIABLE WEEKDAY TO THE NEW DATASET
    new_dataset$weekday <- weekdays(new_dataset$date)
    new_dataset$Type_of_Day[new_dataset$weekday == "Monday" 
                        | new_dataset$weekday == "Tuesday" 
                        | new_dataset$weekday == "Wednesday" 
                        | new_dataset$weekday == "Thursday" 
                        | new_dataset$weekday == "Friday"] <- "WEEKDAY"

    new_dataset$Type_of_Day[new_dataset$weekday == "Saturday" 
                        | new_dataset$weekday == "Sunday"] <- "WEEKEND"
    new_dataset$Type_of_Day <- as.factor(new_dataset$Type_of_Day) 
```
```{r results='hide'} 
  ### AGGREGATE THE TOTAL STEPS GROUPED BY WEEKDAY / WEEKEND
    new_dataset <- data.table(new_dataset)
    new_dataset[, total_steps_per_interval_type_of_day:=sum(steps, na.rm=TRUE), by=c('interval','Type_of_Day')]
  	aggregate_per_interval_type_of_day <- unique(new_dataset[,c('interval','Type_of_Day','total_steps_per_interval_type_of_day')])
```
```{r message = FALSE}
  ##SET THE NUMBER OF PLOTS
    par(mfrow=c(1,1))
  ### DRAW A TIME SERIES PLOT SPLITTED ACROSS VARIOUS INTERVALS
    plot <- ggplot(data = aggregate_per_interval_type_of_day, aes(x=interval, y=total_steps_per_interval_type_of_day)) 
    plot <- plot + xlab("Interval") + ylab("Total Number of Steps") + ggtitle("Number of Steps Grouped by Interval Across Type of Day") 
    plot <- plot + geom_line(color = "#FF0000")
    plot <- plot + facet_grid(Type_of_Day ~ .)
    plot <- plot + theme(plot.title = element_text(color="#666666", face="bold", size=16)) 
    plot <- plot + theme(axis.title = element_text(color="#666666", face="bold", size=14))     
```
```{r results='hide', echo = FALSE} 
    plot
    with(plot,dev.copy(png, 'plot4.png'))
    dev.off()
```
```{r eval = FALSE} 
  ### COPY PLOT TO PLOT4.PNG
    with(plot,dev.copy(png, 'plot4.png'))
  ### CLOSE THE GRAPHICS DEVICE
    dev.off()
```
---
title: "End of Peer Assessment 1"
---